Closure-Free Functional Programming in a Two-Level Type Theory

### 2024-06-12 newpaper/paper.tex
wip

Modern functional programming supports many convenient abstractions. These often
come with significant runtime overheads. Sometimes the overheads are acceptable,
but in other cases compiler optimization is crucial. Monads in Haskell is an
example for the latter. Even the $\Reader$ monad, which is one of the simplest
in terms of implementation, yields large overheads when compiled without
optimizations. Consider the following:
\begin{alignat*}{3}
  &\vf :: \Int \to \Reader\,\Bool\,\Int \\
  &\vf\,\vx = \mdo\{\vb \fro \msf{ask}; \bif \vx\, \bthen \return (\vx + 10)\, \belse \return (\vx + 20)\}
\end{alignat*}
With optimizations enabled, GHC compiles this roughly to the code below:
\begin{alignat*}{3}
  &\vf :: \Int \to \Bool \to \Int \\
  &\vf = \lam \vx\,\vb.\, \bif \vb\, \bthen \vx + 10\, \belse \vx + 20
\end{alignat*}
Without optimizations we roughly get:
\begin{alignat*}{3}
  &\vf = \lam \vx.\,(\bind)\,\msf{monadReader}\,(\msf{ask}\,\msf{monadReaderReader})\,(\lam \vb.\,\bif \vb\\
  &\ind \bthen\return\,\msf{monadReader}\,(\vx + 10)\\
  &\ind \belse\hspace{0.25em}\return\,\msf{monadReader}\,(\vx + 20))
\end{alignat*}
Here, $\msf{monadReader}$ and $\msf{monadReaderReader}$ are runtime
dictionaries, respectively for the $\msf{Monad}$ and $\msf{MonadReader}$
instances, and, for example, $(\bind)\,\msf{monadReader}$ is a field projection
from the dictionary. We get a runtime closure from the $\lam \vb.\,...$
function, and $(\bind)$, $\msf{ask}$ and $\mbf{return}$ also produce additional
closures.

@@

@@

@@

\subsubsection{$\State$ and $\Reader$ operations} If we use
the $\modify$ function that we already have in $\State_\M$, a curious thing
happens. The meaning of $\modify\,(\lam \vx.\,\qt{\spl \vx + \spl \vx})$ is to
replace the current state $\vx$, as an object expression, with the expression
$\qt{\spl \vx + \spl \vx}$, and this happens at unstaging time. This behaves as an
``inline'' modification which replaces every subsequent mention of the state
with a different expression. For instance, ignoring newtype wrappers for now,
\[ \mdown \dlr \mdo \{\modify\,(\lam \vx.\,\qt{\spl \vx + \spl \vx}); \modify\,(\lam \vx.\,\qt{\spl \vx + \spl \vx});\return\,\qt{()}\} \]
is unstaged to
\[ \qt{\lam \vx.\,((),\,(\vx + \vx) + (\vx + \vx))} \]
which duplicates the evaluation of $\vx + \vx$. The duplication can be avoided
by let-binding the result in the object language. A similar phenomenon happens
with the $\mit{local}$ function in $\Reader$. So we define ``stricter'' versions
of these operations. We also return $\Up ()$ from actions instead of $()$ ---
the former is more convenient, because the $\mdown$ operation can be immediately
used on it.
\begingroup
\allowdisplaybreaks
\begin{alignat*}{3}
  & \mput' : (\MonadState\,(\Up \vS)\,\vM,\,\MonadGen\,\vM) \RA \Up \vS \to \vM\,(\Up ()) \\
  & \mput'\,\vs = \mdo \{\vs \fro \gen\,\vs; \mput\,\vs; \return \qt{()}\}\\
  &\\
  & \modify' : (\MonadState\,(\Up \vS)\,\vM,\,\MonadGen\,\vM) \RA (\Up \vS \to \Up \vS) \to \vM\,(\Up ()) \\
  & \modify'\,\vf = \mdo \{\vs \fro \get; \mput'\,(\vf\,\vs)\}\\
  &\\
  & \mit{local'} : (\MonadReader\,(\Up \vR)\,\vM,\,\MonadGen\,\vM) \RA (\Up \vR \to \Up \vR) \to \vM\,\vA \to \vM\,\vA\\
  & \mit{local'}\,\vf\,\vma = \mdo \{\vr \fro \mit{ask}; \vr \fro \gen\,(\vf\,\vr);\mit{local}\,(\lam \_.\,\vr)\,\vma\}
\end{alignat*}
\endgroup
Now,
\[ \mdown \dlr \mdo \{\modify'\,(\lam \vx.\,\qt{\spl \vx + \spl \vx}); \modify'\,(\lam \vx.\,\qt{\spl \vx + \spl \vx})\} \]
is unstaged to
\[ \qt{\lam \vx.\,\letdef x := \vx + \vx; \letdef \vx := \vx + \vx; ((),\,\vx)} \]

@@

\ref{sec:stream-fusion}, so we develop SOP-s. We view $\SOP$-s as a Tarski-style

@@

Now, we can modify our running example as follows.

@@

Then, we can show using the generativity axiom that $\Sigma_\SOP$ supports
projections, pairing and the $\beta\eta$-rules. These are all needed when we
define the $\IsSOP$ instance, when we have to prove that encoding via $\rep$
is an isomorphism. Concretely, assuming $\IsSOP\,\vA$ and $\{\va : \vA\} \to
\IsSOP\,(\vB\,\va)$, we define $\Rep$ and $\msf{rep}$ for $\Sigma\,\vA\,\vB$ as follows:
\[ \Rep\,\{\Sigma\,\vA\,\vB\} = \Sigma\,(\Rep\,\{\vA\})\,(\lam \vx.\,\Rep\,\{\vB\,(\rep^{-1}\,\vx)\}) \]
\[  \rep\,\{\Sigma\,\vA\,\vB\}\,(\va,\,\vb) = (\rep\,\{\vA\}\,\va,\,\rep\,\{\vB\,(\rep^{-1}\,(\rep\,\vx))\}\,\vb) \]
Note that $\rep$ is only well-typed up to the fact that $\rep^{-1} \circ \rep =
\id$; the Agda definition contains an additional transport that we omit here.
This is, in fact, our reason for including the isomorphism equations in
$\IsSOP$. This, in turn, necessitates using SOP instead of finite sums. We can
define a product type former for finite sums of value types, by taking the
pairwise products of components. However, we can only take the
\emph{object-level} products here, and since object-level products have no
$\beta\eta$-rules, we cannot prove $\beta\eta$ for the derived product type in
finite sums, and likewise for the derived $\Sigma$-type. When we use SOP
instead, we do not have to use object-level products and this issue does not
appear.

@@

\subsubsection*{Two-level type theory.} Two-level lambda calculi were first developed in
the context of abstract interpretation and partial evaluation
\cite{DBLP:phd/ethos/Nielson84a,DBLP:books/daglib/0071307}.  This line of
research is characterized by simple types, having the same language features at
different stages, and an emphasis on \emph{binding time analysis},
i.e.\ automatically inferring stage annotations as part of a pipeline for
partial evaluation and program optimization.

@@

We build on both traditions. CFTT itself is a 2LTT and makes heavy use of
dependent type theory. However, we are mostly concerned with code generation and
optimization and borrow concepts from the partial evaluation literature.

@@

\subsubsection*{Sums-of-products} $\SOP$ was proposed for generic Haskell programming
by De Vries and Löh \cite{DBLP:conf/icfp/VriesL14}. Our own $\SOP$
implementation in Haskell is mostly the same as in ibid. The
$\SOP$-of-object-code representation appeared as well in typed Template Haskell
in \cite{DBLP:conf/haskell/PickeringLW20}.

\subsubsection*{CPS and binding-time improvement}
Our $\Gen$ monad can be viewed as a variation on the continuation-passing-style
that has been often used in staged programming. Flanagan et al.'s ANF
translation algorithm uses let-insertion in CPS that works the same way as our
$\gen$ \cite{DBLP:conf/pldi/FlanaganSDF93}. Kiselyov et al.\ likewise uses
essentially the same CPS, in MetaOCaml
\cite{DBLP:conf/popl/KiselyovBPS17}. Jones et al.\ discuss CPS and binding-time
improvement in partial evaluation \cite{partial-evaluation}. $\eta$-expansion
for object-level finite sums is known as ``the trick'' \cite{eta-expansion-trick}.

@@

\subsubsection*{Stream fusion}
The staged stream fusion library $\mathit{strymonas}$ by Kiselyov et al.\ is the
primary prior art \cite{DBLP:conf/popl/KiselyovBPS17}. Here, streams are represented as a sum type of
push and pull representations, allowing both zipping and
$\concatMap$-ing. However, fusion is not guaranteed for all combinations;
zipping two $\concatMap$-s reifies one of the streams in a runtime closure. In
newer versions of the library, fusion is reported to be complete and guaranteed
\cite{DBLP:conf/pepm/0001K24}, however, an exposition of this has not yet been published. Another
notable difference is that $\mathit{strymonas}$ relies on mutable references in
the object language, while we do not.

@@

Coutts developed pull stream fusion in depth in
\cite{DBLP:phd/ethos/Coutts11}. We borrowed the basic design and the basic
stream combinator definitions from there. This account is also close to existing
stream implementations in Haskell. Ibid.\ characterizes fusibility in a
non-staged setting, in terms of ``good consumers'' and ``good producers''.  This
scheme does not cover $\concatMap$.

@@

We give an algebraic \cite{TODO} specification for the models of CFTT and define
the syntax of CFTT to be the initial model. In the following, all type and term
formers are implicitly assumed to be stable under substitution.

@@

\begin{itemize}
\item A category with a terminal object; in the syntax this is the category
  of contexts and parallel substitutions. We use $\Con : \Set$ for contexts,
  $\Sub : \Con \to \Con \to \Set$ for morphisms and $\emptycon : \Con$ for the
  terminal object (which is the empty context in the syntax).
\item A family structure in the sense of category-with-families \cite{TODO},
  specifying metatypes and terms. This structure is additionally indexed by
  universe levels, so we have $\MTy_i : \Con \to \Set$ and $\msf{MetaTm}_i :
  (\Gamma : \Con) \to \MTy_i\,\Gamma \to \Set$ for $i \in \mbb{N}$. We support
  basic type formers, identity types and W-types. We write $\Gamma \ext A$ for
  an extended context. Additionally, we have Coquand-style universes,
  i.e.\ $\U_i : \MTy_{i+1}\,\Gamma$ such that there is an isomorphism
  $\msf{MetaTm}_{i+1}\,\Gamma\,\U_i \simeq \MTy_i\,\Gamma$.  We use $\El$ and
  $\code$ to denote the components of this isomorphism.
\item
  For object-level types, we have $\Ty : \MTy_0\,\Gamma$, $\VTy :
  \MTy_0\,\Gamma$ and $\CTy : \MTy_0\,\Gamma$, together with all object-level
  type formers and inclusion operations $\msf{V} : \MTm\,\Gamma\,\VTy \to \MTm\,\Gamma\,\Ty$
  and $\msf{C} : \MTm\,\Gamma\,\CTy \to \MTm\,\Gamma\,\Ty$.
\item For object-level terms, we have $\Tm : \MTm\,\Gamma\,\Ty \to \Set$,
  supporting context extension, i.e.\ we have $\Gamma,\ext\,A$ for
  $A : \MTm\,\Gamma\,\Ty$ such that $\Sub\,\Gamma\,(\Delta\,\ext\,A)$ is isomorphic
  to $(\sigma : \Sub\,\Gamma\,\Delta) \times \Tm\,\Gamma\,(A[\sigma])$. All
  object-level term formers are supported in $\Tm$.
\item Staging operations: we have $\Lift : \MTm\,\Gamma\,\Ty \to \MTy\,\Gamma$ together
  with isomorphisms $\Tm\,\Gamma\,A \simeq \MTm\,\Gamma\,(\Lift A)$.
  We write $\ql\!\blank\!\qr$ and $\spl\!\blank$ for the components of the isomorphism.
\end{itemize}
\end{definition}
This presentation is a bit more verbose than the surface syntax used in the
paper. First, the meta level uses explicit coding and decoding operations for
universes, while the surface syntax assumes Russell-style universes. Second, the
object level has $\VTy$ and $\CTy$ as Tarski-style universes with explicit
inclusion into $\Ty$, while the surface syntax used implicit coercions.

Also, this is not the only possible presentation. Here, we follow the style of
the informal syntax by having a separate \emph{sort} for object-level terms,
together with explicit quoting and splicing. However, we do not have a sort for
object types and instead specify them as terms of metatypes. This is a mixture
of two styles:
\begin{itemize}
\item The ``classic'' 2LTT style would have separate sorts and staging
  operations for all of $\Ty$, $\CTy$ and $\VTy$. Here the two levels are
  specified independently and level-mixing is only possible through staging
  operations.
\item The ``higher-order abstract syntax'' (HOAS) or ``logical framework'' style
  would only have metatypes and metaterms, no quoting or splicing, and would
  reuse metafunctions to represent object-level binders.Our Agda implementation
  actually follows this style.
\end{itemize}
In the context of CFTT, the choice between HOAS and 2LTT styles is a matter of
taste, and the two are inter-derivable \cite{TODO}.

@@

\subsection{W-types}
So far we have stayed close to \cite{staged2ltt}, with only cosmetic
changes. Supporting W-types, as a way to get inductive families \cite{TODO}, is
the biggest extension here. We extend Section 5 of \cite{staged2ltt} here. In
the syntax of CFTT, $\vW$-types are as follows. The formal specification of
syntactic $\vW{-}\msf{elim}$ is rather gnarly, because of the usage of explicit
substitutions, but we write it out here anyways.

@@

In this section we assume an \emph{extensional} identity type in CFTT.  As we
mentioned in Section \ref{TODO}, we can show soundness of staging with
extensional identity and certain \emph{postulated} identities, but we cannot do
the same with intensional identity. Hence, we formulate unstaging as consisting
of two steps: first, a translation of intensional identity to extensional (which
amounts to proof erasure), followed by evaluation in the presheaf model.

@@

The generativity axiom from Section \ref{TODO} has the following type, in internal
CFTT syntax:
\[ (A : \U_\msf{P})(f : \Elprod\,A \to \USOP)(x\,y : \Elprod\,A) \to f\,x = f\,y  \]
In the presheaf model, we have the following:
\begin{itemize}
\item $\wh{\Uprod}$ is the set (constant presheaf) of lists of object-theoretic types.
\item $\wh{\USOP}$ is the set of lists of lists of object-theoretic types.
\item $\wh{\Elprod\,A}$ is the presheaf such that $\wh{\Elprod\,A}\,\Gamma$ is
  the set of products of object-theoretic terms in context $\Gamma$, with types
  of components taken from $A$.
\end{itemize}
Unfolding the type of generativity in the presheaf model, it suffices to show
that assuming
\begin{itemize}
\item $A : \wh{\Uprod}$, $\Gamma : \Cono$, $x,\,y : \wh{\Elprod\,A}\,\Gamma$
\item and $f : (\Delta : \Cono) \to \Subo\,\Delta\,\Gamma \to \wh{\Elprod\,A}\,\Delta \to \wh{\USOP}$ such that $f$ is natural with respect to object-theoretic substitution,
\end{itemize}
we have that $f\,\Gamma\,\id\,x = f\,\Gamma\,\id\,y$.

>>>>

Modern functional programming supports many convenient abstractions. These often
come with significant runtime overheads. Sometimes the overheads are acceptable,
but in other cases compiler optimization is crucial. Monads in Haskell is an
example for the latter. Even the $\Reader$ monad, which is one of the simplest
in terms of implementation, yields large overheads when compiled without
optimizations. Consider the following:
\begin{alignat*}{3}
  &\vf :: \Int \to \Reader\,\Bool\,\Int \\
  &\vf\,\vx = \mdo\{\vb \fro \msf{ask}; \bif \vx\, \bthen \return (\vx + 10)\, \belse \return (\vx + 20)\}
\end{alignat*}
With optimizations enabled, GHC compiles this roughly to the code below:
\begin{alignat*}{3}
  &\vf :: \Int \to \Bool \to \Int \\
  &\vf = \lam \vx\,\vb.\, \bif \vb\, \bthen \vx + 10\, \belse \vx + 20
\end{alignat*}
Without optimizations we roughly get:
\begin{alignat*}{3}
  &\vf = \lam \vx.\,(\bind)\,\msf{monadReader}\,(\msf{ask}\,\msf{monadReaderReader})\,(\lam \vb.\,\bif \vb\\
  &\ind \bthen\return\,\msf{monadReader}\,(\vx + 10)\\
  &\ind \belse\hspace{0.25em}\return\,\msf{monadReader}\,(\vx + 20))
\end{alignat*}
Here, $\msf{monadReader}$ and $\msf{monadReaderReader}$ are runtime
dictionaries, respectively for the $\msf{Monad}$ and $\msf{MonadReader}$
instances, and, for example, $(\bind)\,\msf{monadReader}$ is a field projection
from the dictionary. This results from the dictionary-passing elaboration of
type classes \cite{DBLP:conf/popl/WadlerB89}. We get a runtime closure from the
$\lam \vb.\,...$ function, and $(\bind)$, $\msf{ask}$ and $\mbf{return}$ also
produce additional closures.

@@

Why not just make the object language less liberal, e.g.\ by disallowing
$\lambda$ under $\mbf{case}$ or $\mbf{let}$, thereby making call saturation
easier or more obvious? There is a trade-off between making the object language
more restricted, and thus easier to compile, and making \emph{metaprogramming}
more convenient. We will see that the ability to insert $\mbf{let}$-s without
restriction is very convenient in code generation, and likewise the ability to
have arbitrary object expressions in $\mbf{case}$ bodies. In this paper we go
with the most liberal object syntax, at the cost of needing more downstream
processing.

On the other side of the spectrum, one might imagine generating object code in
low-level A-normal form. This is more laborious, but it could be also
interesting, because it forces us to invent abstractions for manipulating ANF in
the metalanguage.

@@

%% On the other side of the spectrum, one might imagine generating object code in
%% low-level A-normal form. This is more laborious, but it could be also
%% interesting, because it forces us to invent abstractions for manipulating ANF in
%% the metalanguage. In a similar vein, Allais recently proposed metaprogramming
%% quantum circuits in a 2LTT \cite{TODO}. We leave such setups
%% with low-level object languages to future investigation.

@@

\subsubsection{$\State$ and $\Reader$ operations} If we use
the $\modify$ function that we already have in $\State_\M$, a curious thing
happens. The meaning of $\modify\,(\lam \vx.\,\qt{\spl \vx + \spl \vx})$ is to
replace the current state $\vx$, as an object expression, with the expression
$\qt{\spl \vx + \spl \vx}$, and this happens at unstaging time. This behaves as an
``inline'' modification which replaces every subsequent mention of the state
with a different expression. For instance, ignoring newtype wrappers for now,
\[ \mdown \dlr \mdo \{\modify\,(\lam \vx.\,\qt{\spl \vx + \spl \vx}); \modify\,(\lam \vx.\,\qt{\spl \vx + \spl \vx});\return\,\qt{()}\} \]
is unstaged to
\[ \qt{\lam \vx.\,((),\,(\vx + \vx) + (\vx + \vx))} \]
which duplicates the evaluation of $\vx + \vx$. The duplication can be avoided
by let-binding the result in the object language. A similar phenomenon happens
with the $\mit{local}$ function in $\Reader$. So we define ``stricter'' versions
of these operations. We also return $\Up ()$ from actions instead of $()$ ---
the former is more convenient, because the $\mdown$ operation can be immediately
used on it.
\begingroup
\allowdisplaybreaks
\begin{alignat*}{3}
  & \mput' : (\MonadState\,(\Up \vS)\,\vM,\,\MonadGen\,\vM) \RA \Up \vS \to \vM\,(\Up ()) \\
  & \mput'\,\vs = \mdo \{\vs \fro \gen\,\vs; \mput\,\vs; \return \qt{()}\}\\
  &\\
  & \modify' : (\MonadState\,(\Up \vS)\,\vM,\,\MonadGen\,\vM) \RA (\Up \vS \to \Up \vS) \to \vM\,(\Up ()) \\
  & \modify'\,\vf = \mdo \{\vs \fro \get; \mput'\,(\vf\,\vs)\}\\
  &\\
  & \mit{local'} : (\MonadReader\,(\Up \vR)\,\vM,\,\MonadGen\,\vM) \RA (\Up \vR \to \Up \vR) \to \vM\,\vA \to \vM\,\vA\\
  & \mit{local'}\,\vf\,\vma = \mdo \{\vr \fro \mit{ask}; \vr \fro \gen\,(\vf\,\vr);\mit{local}\,(\lam \_.\,\vr)\,\vma\}
\end{alignat*}
\endgroup
Now,
\[ \mdown \dlr \mdo \{\modify'\,(\lam \vx.\,\qt{\spl \vx + \spl \vx}); \modify'\,(\lam \vx.\,\qt{\spl \vx + \spl \vx})\} \]
is unstaged to
\[ \qt{\lam \vx.\,\letdef x := \vx + \vx; \letdef \vx := \vx + \vx; ((),\,\vx)}. \]

@@

\ref{sec:stream-fusion}, so we develop SOP-s. SOPs have been used in generic and
staged programming; see e.g.\ \cite{sop} and
\cite{DBLP:conf/haskell/PickeringLW20}. We view $\SOP$-s as a Tarski-style

@@

Now, the following definition yields the previously seen code output with
$\msf{joinNothing}$ and $\msf{joinJust}$.

@@

Then, we can show using the generativity axiom that $\Sigma_\SOP$ supports
projections, pairing and the $\beta\eta$-rules. These are all needed when we
define the $\IsSOP$ instance, when we have to prove that encoding via $\rep$
is an isomorphism. Concretely, assuming $\IsSOP\,\vA$ and $\{\va : \vA\} \to
\IsSOP\,(\vB\,\va)$, we define $\Rep$ and $\msf{rep}$ for $\Sigma\,\vA\,\vB$ as follows:
\[ \Rep\,\{\Sigma\,\vA\,\vB\} = \Sigma\,(\Rep\,\{\vA\})\,(\lam \vx.\,\Rep\,\{\vB\,(\rep^{-1}\,\vx)\}) \]
\[  \rep\,\{\Sigma\,\vA\,\vB\}\,(\va,\,\vb) = (\rep\,\{\vA\}\,\va,\,\rep\,\{\vB\,(\rep^{-1}\,(\rep\,\vx))\}\,\vb) \]
Note that $\rep$ is only well-typed up to the fact that $\rep^{-1} \circ \rep =
\id$; the Agda definition contains an additional transport that we omit here.
This is, in fact, our reason for including the isomorphism equations in
$\IsSOP$.

This, in turn, necessitates using SOP instead of finite sums, for the following
reason. We can define a product type former for finite sums of value types, by
taking the pairwise products of components, but in this case the products would
be object-level products, which do not support $\beta\eta$ rules. This implies
that we cannot prove $\beta\eta$-rules for the derived product type in finite
sums. In contrast, when we define products for SOP-s, we take \emph{meta-level}
pairwise products of components, which does support $\beta$ and $\eta$.

@@

\subsubsection*{Two-level calculi.}
Two-level lambda calculi were first developed in the context of abstract
interpretation and partial evaluation
\cite{DBLP:phd/ethos/Nielson84a,DBLP:books/daglib/0071307}.  This line of
research is characterized by simple types, having the same language features at
different stages, and an emphasis on \emph{binding time analysis},
i.e.\ automatically inferring stage annotations as part of a pipeline for
partial evaluation and program optimization.

@@

\subsubsection*{Higher-order abstract syntax and logical frameworks.}
Hofmann's work on the semantics of higher-order abstract syntax (HOAS) is an
important precursor to our work \cite{Hofmann:1999:SAH:788021.788940}. It
anticipates many of the later developments in logical frameworks and two-level
type theories. In particular, we build on Hofmann's presheaf model for our 2LTT
semantics, and his sketch of adequacy for HOAS corresponds to our definition of
unstaging and its soundness proof. In general, there is a close correspondence
between logical frameworks that represent object languages in HOAS style
(e.g.\ \cite{Harper93lf}) and two-level type theories, and in some cases the two
kinds of presentations are inter-derivable \cite[Section
  3.3]{DBLP:journals/corr/abs-2302-08837}.

Cocon is a logical framework that has a dependently typed meta level
(``computation'' level) and an object language that is configured by a signature
\cite{DBLP:journals/corr/abs-1901-03378}. Cocon, like 2LTTs, can be used as a
dependently typed metaprogramming language, but there are several
differences. First, Cocon supports and emphasizes intensional analysis
(i.e.\ the ability to do structural induction on object code), which is made
possible through a contextual modality \cite{nanevski2008contextual}. In
contrast, intensional analysis is missing from 2LTTs. However, in Cocon, we can
only metaprogram object-level terms by specifying their object-level contexts
(i.e.\ possible free variables), while in 2LTTs the object-level contexts are
implicit and are only computed during code generation. The latter style is
more convenient for staged compilation applications, where intensional analysis
is rarely used and explicit context tracking would add a significant amount of
bureaucracy.

\subsubsection*{CPS and monads in staged programming} Writing code generators
in continuation-passing style was first proposed by Bondorf
\cite{DBLP:conf/lfp/Bondorf92}. Flanagan et al.'s ANF translation algorithm uses
CPS for let-insertion as well \cite{DBLP:conf/pldi/FlanaganSDF93}, similarly to
our $\gen$ function. These sources do no use explicit monadic notation
though. Jones et al.\ discuss CPS and binding-time improvement in partial
evaluation in \cite{partial-evaluation}. In
\cite{DBLP:conf/emsoft/KiselyovST04}, \cite{DBLP:conf/pepm/SwadiTKP06} and
\cite{DBLP:journals/scp/CaretteK11}, the composition of state and continuation
monads was used, similarly to our $\StateT\,S\,\Gen$, but without polymorphic
answer types. The definition of let-insertion here is again similar to ours.

@@

\subsubsection*{Sums-of-products} $\SOP$ was proposed for generic Haskell programming
by De Vries and Löh \cite{DBLP:conf/icfp/VriesL14}. Our own $\SOP$
implementation in Haskell is mostly the same as in ibid. The
$\SOP$-of-object-code representation appeared as well in typed Template Haskell
in \cite{DBLP:conf/haskell/PickeringLW20}. $\eta$-expansion for object-level
finite sums is known as ``the trick'' \cite{eta-expansion-trick}.

@@

\subsubsection*{Stream fusion}
The staged stream fusion library $\mathit{strymonas}$ by Kiselyov et al.\ is the
primary prior art \cite{DBLP:conf/popl/KiselyovBPS17}. Here, streams are
represented as a sum type of push and pull representations, allowing both
zipping and $\concatMap$-ing. However, fusion is not guaranteed for all
combinations; zipping two $\concatMap$-s reifies one of the streams in a runtime
closure. In newer versions of the library, fusion is complete
\cite{DBLP:conf/pepm/0001K24}, however, an exposition of this has not yet been
published. Additionally, $\mathit{strymonas}$ heavily relies on mutable
references in the object language, and also uses some switching on object-level
data to implement control flow. This style of code output can be still reliably
optimized by downstream compilers (C or OCaml), but our solution is more
conservative in that it does not use any mutation or runtime switching.
This ``purity'' of our streams also makes them easier to generalize, e.g.\
by using arbitrary $\MonadGen$ monads instead of $\Gen$ in stream internals.

@@

Coutts developed pull stream fusion in depth in
\cite{DBLP:phd/ethos/Coutts11}. We borrowed the basic design and the basic
stream combinator definitions from there. This account is also close to existing
stream implementations in Haskell. Coutts characterized fusibility in a
non-staged setting, in terms of ``good consumers'' and ``good producers''.  This
scheme does not cover $\concatMap$.

@@

We give an algebraic \cite{cartmellthesis,DBLP:journals/corr/abs-2302-08837}
specification for the models of CFTT and define the syntax of CFTT to be the
initial model. In the following, all type and term formers are implicitly
assumed to be stable under substitution.

@@

\begin{itemize}
\item A category with a terminal object; in the syntax this is the category
  of contexts and parallel substitutions. We use $\Con : \Set$ for contexts,
  $\Sub : \Con \to \Con \to \Set$ for morphisms and $\emptycon : \Con$ for the
  terminal object (which is the empty context in the syntax).
\item A family structure in the sense of category-with-families \cite{Dybjer96internaltype},
  specifying metatypes and terms. This structure is additionally indexed by
  universe levels, so we have $\MTy_i : \Con \to \Set$ and $\msf{MetaTm}_i :
  (\Gamma : \Con) \to \MTy_i\,\Gamma \to \Set$ for $i \in \mbb{N}$. We support
  basic type formers, identity types and W-types. We write $\Gamma \ext A$ for
  an extended context. Additionally, we have Coquand-style universes,
  i.e.\ $\U_i : \MTy_{i+1}\,\Gamma$ such that there is an isomorphism
  $\msf{MetaTm}_{i+1}\,\Gamma\,\U_i \simeq \MTy_i\,\Gamma$.  We use $\El$ and
  $\code$ to denote the components of this isomorphism.
\item
  For object-level types, we have $\Ty : \MTy_0\,\Gamma$, $\VTy :
  \MTy_0\,\Gamma$ and $\CTy : \MTy_0\,\Gamma$, together with all object-level
  type formers and inclusion operations $\msf{V} : \MTm\,\Gamma\,\VTy \to \MTm\,\Gamma\,\Ty$
  and $\msf{C} : \MTm\,\Gamma\,\CTy \to \MTm\,\Gamma\,\Ty$.
\item For object-level terms, we have $\Tm : \MTm\,\Gamma\,\Ty \to \Set$,
  supporting context extension, i.e.\ we have $\Gamma,\ext\,A$ for
  $A : \MTm\,\Gamma\,\Ty$ such that $\Sub\,\Gamma\,(\Delta\,\ext\,A)$ is isomorphic
  to $(\sigma : \Sub\,\Gamma\,\Delta) \times \Tm\,\Gamma\,(A[\sigma])$. All
  object-level term formers are supported in $\Tm$.
\item Staging operations: we have $\Lift : \MTm\,\Gamma\,\Ty \to \MTy\,\Gamma$ together
  with isomorphisms $\Tm\,\Gamma\,A \simeq \MTm\,\Gamma\,(\Lift A)$.
  We write $\ql\!\blank\!\qr$ and $\spl\!\blank$ for the components of the isomorphism.
\end{itemize}
\end{definition}
This presentation is a bit more verbose than the surface syntax used in the
paper. First, the meta level uses explicit coding and decoding operations for
universes, while the surface syntax assumes Russell-style universes. Second, the
object level has $\VTy$ and $\CTy$ as Tarski-style universes with explicit
inclusion into $\Ty$, while the surface syntax used implicit coercions.

Also, this is not the only possible presentation. Here, we follow the style of
the informal syntax by having a separate \emph{sort} for object-level terms,
together with explicit quoting and splicing. However, we do not have a sort for
object types and instead specify them as terms of metatypes. This is a mixture
of two styles:
\begin{itemize}
\item The ``classic'' 2LTT style would have separate sorts and staging
  operations for all of $\Ty$, $\CTy$ and $\VTy$. Here the two levels are
  specified independently and level-mixing is only possible through staging
  operations.
\item The ``higher-order abstract syntax'' (HOAS) or ``logical framework'' style
  would only have metatypes and metaterms, no quoting or splicing, and would
  reuse metafunctions to represent object-level binders.Our Agda implementation
  actually follows this style.
\end{itemize}
In the context of CFTT, the choice between HOAS and 2LTT styles is a matter of
taste, and the two are inter-derivable \cite[Section
  3.3]{DBLP:journals/corr/abs-2302-08837}.

@@

\subsection{W-types}
So far we have stayed close to \cite{staged2ltt}, with only cosmetic
changes. Supporting W-types, as a way to get inductive families
\cite{inductivefamilies}, is the biggest extension here. We extend Section 5 of
\cite{staged2ltt} here. In the syntax of CFTT, $\vW$-types are as follows. The
formal specification of syntactic $\vW{-}\msf{elim}$ is rather gnarly, because
of the usage of explicit substitutions, but we write it out here anyways.

@@

In this section we assume an \emph{extensional} identity type in CFTT. As we
mentioned in Section \ref{sec:semantics-of-staging}, we can show soundness of staging with
extensional identity and certain \emph{postulated} identities, but we cannot do
the same with intensional identity. Hence, we formulate unstaging as consisting
of two steps: first, a translation of intensional identity to extensional (which
amounts to proof erasure), followed by evaluation in the presheaf model.

@@

The generativity axiom from Section \ref{sec:concatmap-for-streams} has the following type, in internal
CFTT syntax:
\[ (A : \U_\msf{P})(f : \Elprod\,A \to \USOP)(x\,y : \Elprod\,A) \to f\,x = f\,y  \]
In the presheaf model, we have the following:
\begin{itemize}
\item $\wh{\Uprod}$ is the set (constant presheaf) of lists of object-theoretic types.
\item $\wh{\USOP}$ is the set of lists of lists of object-theoretic types.
\item $\wh{\Elprod\,A}$ is the presheaf such that $\wh{\Elprod\,A}\,\Gamma$ is
  the set of products of object-theoretic terms in context $\Gamma$, with types
  of components taken from $A$.
\end{itemize}
Unfolding the type of generativity in the presheaf model, it suffices to show
that assuming
\begin{itemize}
\item $A : \wh{\Uprod}$, $\Gamma : \Cono$, $x,\,y : \wh{\Elprod\,A}\,\Gamma$
\item and $f : (\Delta : \Cono) \to \Subo\,\Delta\,\Gamma \to \wh{\Elprod\,A}\,\Delta \to \wh{\USOP}$ such that $f$ is natural with respect to object-theoretic substitution,
\end{itemize}
we have that $f\,\Gamma\,\id\,x = f\,\Gamma\,\id\,y$.
